\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-26 paper.tex
Clear the repo

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, enhanced, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi

@@

Hexagonating NLDR data (see Figure~\ref{fig-meth} Step 2) involves
partitioning the 2D embedding data into hexagonal bins, a technique
commonly referred to as hexagonal binning
\citep[\citet{article66}]{Carr1987}. This method use a hexagonal grid to
create a bivariate histogram that effectively visualizes the structure
of high-D data. Hexagons, one of only three regular polygons capable of
tessellating a plane \citep{Carr2013}, offer unique advantages due to
their symmetry of nearest neighbors and maximal number of sides for such
tessellations \citep{Dan2023}. This geometric property makes hexagons
more efficient in covering the plane compared to other regular
tessellations and reduces visual bias when displaying data densities
\citep{Dan2023}. In our algorithm, we aim to conduct regular hexagonal
binning (\textbf{cite quollr paper}), involving the computation of
hexagonal grid configurations, the generation of the grid, and the
assignment of the 2D embedding to hexagons.

@@

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the \textbf{UMAP} embeddings of \textbf{S-curve} training data: (a)
b = 12 (3, 4), h = 0.62, (b) b = 84 (6, 14), h = 0.13, and (c) b = 336
(12, 28), h = 0.06. The hexbins are colored based on the density of
points, with darker colors indicating higher point density and yellow
color representing lower point density within each bin. What is the
number of bins that would be effective in representing low-dimensional
data?}

@@

@@

@@

@@

According to \citet{Dan2023}, the hexagonal binning is done by
tessellating the \(xy\) plane over the set (range(\(x\)), range(\(y\)))
(see \textbf{?@fig-scurveshifthexgridsexp} (b)). In that case, bin
centroids are defined as shown in \textbf{?@fig-scurveshifthexgridsexp}
(a) with gray colour. Rather than sticking to the typical hexagonal
grid, introducing a meaningful shift in both the x and y directions
presents an opportunity for an improved 2D model. Therefore,
investigating this shift in the hexagonal grid is an important parameter
to consider.

As shown in \textbf{?@fig-scurveshifthexgrids}, the shifting influences
the distribution of points and number of non-empty bins, impacting the
resulting 2D model. According to
\textbf{?@fig-diagnosticpltScurvehexbins}, the 2D model with a total of
\(144\) bins applied to S-curve UMAP data does not require any shifting
because the lowest MSE occurs when no shift is introduced.

@@

Residuals are essential for evaluating the accuracy of representing
high-D points by the high-D mapping of 2D bin centroids. To measure this
accuracy, an error metric is introduced, quantifying the sum of squared
differences between the high-D data (\(x_{ij}\)) and the high-D mapping
of the 2D bin centroid data (\(C_{x_ij}\)) across all observations and
dimensions (see Equation~\ref{eq-equation11}).

\begin{equation}\protect\hypertarget{eq-equation11}{}{
\text{Error} = \sum_{j = 1}^{n}\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2
}\label{eq-equation11}\end{equation}

@@

Visualizing the models alongside the original high-D data provides
insights into how different techniques capture the underlying clustering
structure. The tSNE model exhibits five well-separated clusters,
effectively preserving both local and global structures (see video link
of \textbf{?@fig-modelfiveGau} (a)). UMAP, on the other hand, also
presents five distinct clusters but with a more flattened surface
appearance (see video link of \textbf{?@fig-modelfiveGau} (b)). The
PHATE model shows five separated clusters resembling triangles or
partial triangles but struggles to capture local structures (see video
link of \textbf{?@fig-modelfiveGau} (c)). TriMAP and PaCMAP both display
five well-separated clusters with flat surfaces, each capturing
within-cluster variation to varying extents. Despite the various 2D
representations, all NLDR techniques preserve the global structure (see
video link of \textbf{?@fig-modelfiveGau} (d), (e) respectively).
However, tSNE, effectively capture both local and global structures, as
indicated by lower AIC values in Figure~\ref{fig-diagnosticpltGau} (a).

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-diagnosticpltGau-1.pdf}

@@

\caption{\label{fig-diagnosticpltGau}Goodness of fit statistics from
different NLDR techniques applied to training five spherical Gaussian
cluster dataset. What is the best NLDR technique to represent the
original data in 2D?}

@@

@@

\caption{\label{fig-umapauthor}(a) 2D layout from UMAP (n\_neighbors =
30, min\_dist = 0.3) applied for the PBMC3k dataset. Is this a best
representation of the original data?}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-msepltpbmc-1.pdf}

\caption{\label{fig-msepltpbmc}Goodness of fit statistics from UMAP
applied to training PBMC3k dataset. What is the effective number of bins
in UMAP to create a 2D model? The MSE plot have a steep slope at the
beginning, indicating that a smaller number of bins causes a larger
amount of error. Then, the slope gradually declines or level off,
indicating that a higher number of bins generates a smaller error. Using
the elbow method, when the total number of bins is set to 588, the slope
of the MSE plot experiences a sudden and noticeable change, resembling
an elbow-like shape. This point indicates that adding less bins does not
enough to capture the data structure.}

@@

\caption{\label{fig-nldervisPBMCUMAP}(a) Model in the 2D space overlaid
on UMAP data, and (b) Error occurred by the high-D model for high-D data
and the error mapped into 2D NLDR points.}

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-72-1.pdf}

@@

The MNIST dataset consists of grayscale images of handwritten digits
\citep{lecun2010}. \citet{Yingfan2021} used this dataset to demonstrate
how the PaCMAP preserves local structure. We select the 2D embedding of
PaCMAP for the handwritten digit 1. As shown in
Figure~\ref{fig-pacmapauthor}, we observe a nonlinear structure present
in 2D space. To answer the following questions,

@@

@@

\caption{\label{fig-sampleimg}Images of the handwritten digit 1 are
distributed along the left-to-right of first embedding axis of the
PaCMAP nonlinear structure.}

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltMNIST-1.pdf}

\caption{\label{fig-diagnosticpltMNIST}Goodness of fit statistics from
PaCMAP applied to training MNIST dataset. What is the effective number
of bins in PaCMAP to create a 2D model? The MSE plot have a steep slope
at the beginning, indicating that a smaller number of bins causes a
larger amount of error. Then, the slope gradually declines or level off,
indicating that a higher number of bins generates a smaller error. Using
the elbow method, when the total number of bins is set to 180, the slope
of the Mean Squared Error (MSE) plot experiences a sudden and noticeable
change, resembling an elbow-like shape. This point indicates that adding
less bins does not enough to capture the data structure.}

@@

\caption{\label{fig-nldervisMNISTPACMAP}(a) Model in the 2D space
overlaid on PaCMAP data, and (b) high-D model error in model space.}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

Hexagonating NLDR data (see Figure~\ref{fig-meth} Step 2) involves
partitioning the NLDR data into hexagonal bins, a technique commonly
referred to as hexagonal binning \citep[\citet{article66}]{Carr1987}.
This method use a hexagonal grid to create a bivariate histogram that
effectively visualizes the structure of high-D data. Hexagons, one of
only three regular polygons capable of tessellating a plane
\citep{Carr2013}, offer unique advantages due to their symmetry of
nearest neighbors and maximal number of sides for such tessellations
\citep{Dan2023}. This geometric property makes hexagons more efficient
in covering the plane compared to other regular tessellations and
reduces visual bias when displaying data densities \citep{Dan2023}.

In our algorithm, we aim to conduct regular hexagonal binning, involving
the computation of hexagonal grid configurations, the generation of the
grid, and the assignment of the NLDR data points to hexagons.

\textbf{(a) Computation of hexagonal grid configurations}

In the computation of hexagonal bin configurations, there are mainly two
configurations to consider: the number of bins along the x and y axes.
The first step involves in determining the hexagonal size (\(s\)) and
the buffer along the x and y axes (\(buffer_{x}\) and \(buffer_{y}\)).
Here, the hexagonal size (\(s\)) represents the radius of the outer
circle surrounding the hexagon. The buffer is important in expanding the
boundary to accommodate potential outliers or edge cases.

When computing the number of bins along the x and y axes (\(b_1\) and
\(b_2\)), the process begins by calculating the range of the respective
2D embedding component. After that, the range is adjusted to accommodate
the buffer amount. Then, the spacing between hexagons is determined
based on \(s\). Finally, the number of bins along the axis is computed
by dividing the adjusted range by the spacing.

For the computation of the number of bins along the x-axis (\(b_1\))
(see Equation~\ref{eq-equation12}), the range of the first embedding
component is defined as \(r_1\), while the buffer amount along the
x-axis is denoted as \(buffer_{x}\), and the horizontal spacing is
represented by \(h\) (see Equation~\ref{eq-equation11}). On the other
hand, the number of bins along the y-axis (\(b_2\)) (see
Equation~\ref{eq-equation14}) is computed based on the range of the
second embedding component (\(r_2\)), the buffer amount along the y-axis
(\(buffer_{y}\)), and the vertical spacing (\(v\)) (see
Equation~\ref{eq-equation13}).

\begin{equation}\protect\hypertarget{eq-equation11}{}{
 h = \sqrt{3} * s
}\label{eq-equation11}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation12}{}{
 b_1 = \frac{r_1 + buffer_{x}}{h}
}\label{eq-equation12}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation13}{}{
  v = 1.5 * s
}\label{eq-equation13}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation14}{}{
 b_2 = \frac{r_2 + buffer_{y}}{v}
}\label{eq-equation14}\end{equation}

\textbf{(b) Generation of the hexagonal grid}

In this process, the first step involves generating centroids within the
hexagonal grid. This begins by defining the starting coordinates for the
hexagons. With the number of bins computed along the x and y axes, along
with the horizontal and vertical spacing, the centroids are iteratively
computed starting from the hexagonal starting coordinates.

Once the centroids for the hexagonal grid are obtained, the next step is
to compute the hexagonal coordinates. For example, if the centroid of a
hexagonal bin is defined as \((C_x, C_y)\), the hexagonal coordinates
can be computed using \(d_x\) (see Equation~\ref{eq-equation15}) and
\(d_y\) (see Equation~\ref{eq-equation16}) (see
Figure~\ref{fig-hexcoord}).

\begin{equation}\protect\hypertarget{eq-equation15}{}{
d_x = \frac{h}{2}
}\label{eq-equation15}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation16}{}{
d_y = \frac{2* v * 1.15}{\sqrt{3}} 
}\label{eq-equation16}\end{equation}

{\centering \includegraphics[width=1\textwidth,height=1\textheight]{figures/hex_coord.png}

\caption{\label{fig-hexcoord}The hexagonal coordinates computed for the
centroid \((C_x, C_y)\).}

\textbf{(c) Assignment of the NLDR data points to hexagons}

After obtaining the centroids of the hexagonal bins for the entire grid,
the process of assigning NLDR points to hexagons involves determining
the nearest hexagonal bin for each NLDR point using the 2D Euclidean
distance. Then, the hexagonal ID of the nearest hexagon is assigned to
the corresponding NLDR points.

@@

\caption{\label{fig-binsize}Hexbin plots from different number of bins
for the \textbf{UMAP} embeddings of \textbf{S-curve} training data: (a)
b = 12 (3, 4), s = 0.62, (b) b = 84 (6, 14), s = 0.13, and (c) b = 336
(12, 28), s = 0.06. The hexbins are colored based on the density of
points, with darker colors indicating higher point density and yellow
color representing lower point density within each bin. What is the
number of bins that would be effective in representing low-dimensional
data?}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-diagnosticpltScurvelwd-1.pdf}

\caption{\label{fig-diagnosticpltScurvelwd}Goodness of fit statistics
from UMAP applied to training S-curve dataset with different benchmark
values to remove the low-density heaxgons. What is the effective
benchmark value to remove the low-density heaxgons? The MSE plot have a
steep slope at the beginning, indicating that a smaller benchmark causes
a small amount of error. Then, the slope gradually increases or level
up, indicating that a higher number of benchmark values generates a
higher error. Using the reverse elbow method, when the benchmark value
is set to 0.2, the slope of the Mean Squared Error (MSE) plot
experiences a sudden and noticeable change, resembling an elbow-like
shape. This point indicates that higher benchmark values remove the
necessary bins as well which lead to distruct the 2D structure.}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-rmlowdenshex-1.pdf}

@@

\caption{\label{fig-rmlowdenshex}Goodness of f}

@@

According to \citet{Dan2023}, the hexagonal binning is done by
tessellating the \(xy\) plane over the set (range(\(x\)), range(\(y\)))
(see Figure~\ref{fig-scurveshifthexgridsexp} (b)). In that case, bin
centroids are defined as shown in
Figure~\ref{fig-scurveshifthexgridsexp} (a) with gray colour. Rather
than sticking to the typical hexagonal grid, introducing a meaningful
shift in both the x and y directions presents an opportunity for an
improved 2D model. Therefore, investigating this shift in the hexagonal
grid is an important parameter to consider.

As shown in Figure~\ref{fig-scurveshifthexgrids}, the shifting
influences the distribution of points and number of non-empty bins,
impacting the resulting 2D model. According to
\textbf{?@fig-diagnosticpltScurvehexbins}, the 2D model with a total of
\(144\) bins applied to S-curve UMAP data does not require any shifting
because the lowest MSE occurs when no shift is introduced.

\includegraphics{paper_files/figure-pdf/unnamed-chunk-36-1.pdf}

{\centering \includegraphics{paper_files/figure-pdf/fig-scurveshifthexgridsexp-1.pdf}

\caption{\label{fig-scurveshifthexgridsexp}(a) Visualization of bin
centroids and (b) hexagonal grid, showcasing the configuration before
(colored in gray) and after (colored in light orange) shifting the
starting point by an identical amount applied in both x and y
directions, with a shift value of \(0.537285\).}

{\centering \includegraphics{paper_files/figure-pdf/fig-scurveshifthexgrids-1.pdf}

\caption{\label{fig-scurveshifthexgrids}Hexbin plots shows the
distribution of points before and after shifting the starting point of
the hexagonal grid generated for UMAP applied to the training S-curve
dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after
shifting (identical shift applied in both x and y directions, with a
shift value of \(0.537285\)).}

@@

Residuals are essential for evaluating the accuracy of representing
high-D points by the high-D mapping of 2D bin centroids. To measure this
accuracy, an error metric is introduced, quantifying the sum of squared
differences between the high-D data (\(x_{ij}\)) and the high-D mapping
of the 2D bin centroid data (\(C_{x_ij}\)) across all observations and
dimensions (see Equation~\ref{eq-equation10}).

\begin{equation}\protect\hypertarget{eq-equation10}{}{
\text{Error} = \sum_{j = 1}^{n}\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2
}\label{eq-equation10}\end{equation}

@@

Visualizing the models alongside the original high-D data provides
insights into how different techniques capture the underlying clustering
structure. The tSNE model exhibits five well-separated clusters,
effectively preserving both local and global structures (see video link
of Figure~\ref{fig-modelfiveGau} (a)). UMAP, on the other hand, also
presents five distinct clusters but with a more flattened surface
appearance (see video link of Figure~\ref{fig-modelfiveGau} (b)). The
PHATE model shows five separated clusters resembling triangles or
partial triangles but struggles to capture local structures (see video
link of Figure~\ref{fig-modelfiveGau} (c)). TriMAP and PaCMAP both
display five well-separated clusters with flat surfaces, each capturing
within-cluster variation to varying extents. Despite the various 2D
representations, all NLDR techniques preserve the global structure (see
video link of Figure~\ref{fig-modelfiveGau} (d), (e) respectively).
However, tSNE, effectively capture both local and global structures.

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-modelfiveGau-1.pdf}

@@

\caption{\label{fig-modelfiveGau}Is there a best model to represent the
original data in 2D space or are they all providing equivalent
information?, (a) Model generated with 21 and 28 bins along the x and y
axes, respectively, using a hexagonal size of 0.03 in the 2D space with
tSNE, (b) Model generated with 60 and 79 bins along the x and y axes,
respectively, using a hexagonal size of 0.01 in the 2D space with UMAP,
(c) Model generated with 60 and 1927 bins along the x and y axes,
respectively, using a hexagonal size of 0.01 in the 2D space with PHATE,
(d) Model generated with 46 and 61 bins along the x and y axes,
respectively, using a hexagonal size of 0.013 in the 2D space with
TriMAP, and (e) Model generated with 31 and 40 bins along the x and y
axes, respectively, using a hexagonal size of 0.02 in the 2D space with
PaCMAP.}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_1.png}

\subcaption{\label{fig-gau1_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_2.png}

\subcaption{\label{fig-gau1_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_3.png}

\subcaption{\label{fig-gau1_sc3}}
\end{minipage}%

\caption{\label{fig-gau1_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/oQxEb4wRdHI}).}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_1.png}

\subcaption{\label{fig-gau2_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_2.png}

\subcaption{\label{fig-gau2_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_3.png}

\subcaption{\label{fig-gau2_sc3}}
\end{minipage}%

\caption{\label{fig-gau2_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/JW49csPpDx4}).}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_1.png}

\subcaption{\label{fig-gau3_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_2.png}

\subcaption{\label{fig-gau3_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_3.png}

\subcaption{\label{fig-gau3_sc3}}
\end{minipage}%

\caption{\label{fig-gau3_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/xqhRODt8Z-o}).}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_1.png}

\subcaption{\label{fig-gau4_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_2.png}

\subcaption{\label{fig-gau4_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_3.png}

\subcaption{\label{fig-gau4_sc3}}
\end{minipage}%

\caption{\label{fig-gau4_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/X-uHSB4VoZ0}).}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_1.png}

\subcaption{\label{fig-gau5_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_2.png}

\subcaption{\label{fig-gau5_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_3.png}

\subcaption{\label{fig-gau5_sc3}}
\end{minipage}%

\caption{\label{fig-gau5_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/UxvU8HQd7JM}).}

@@

In the field of single-cell studies, a common analytical task involves
clustering to identify groups of cells with similar expression profiles.
Analysts often turn to NLDR techniques to verify and identify these
clusters and explore developmental trajectories. To illustrate the
importance of NLDR techniques and parameter selection in identifying
clusters, Human Peripheral Blood Mononclear Cells (PBMC3k) dataset
\citep{chen2023} is used. In a study by \citet{chen2023}, this dataset
was used to demonstrate how UMAP represents clusters (see
Figure~\ref{fig-umapauthor}). As shown in Figure~\ref{fig-umapauthor},
there are three distinct and well-separated clusters.

@@

\caption{\label{fig-umapauthor}2D layout from UMAP (n\_neighbors = 30,
min\_dist = 0.3) applied for the PBMC3k dataset. Is this a best
representation of the original data?}

@@

To determine whether the UMAP representation with the parameter choice
suggested by \citet{chen2023} preserves the original data structure, we
visualize the model constructed with UMAP overlaid on the high-D data
(model-in-data space). The figures in Figure~\ref{fig-pbmc1_sc} show
three well-separated clusters, indicating that the suggested UMAP
representation preserves the global structure (see
Figure~\ref{fig-umapauthor}). However, as shown in
Figure~\ref{fig-pbmc1_sc}, these clusters are close to each other in
high-D space. Also, nonlinear continuity patterns and high-density
patches within the clusters are observed (see
Figure~\ref{fig-pbmc1_sc}). Therefore, the suggested UMAP representation
(see Figure~\ref{fig-umapauthor}) does not accurately preserve the local
structure of the PBMC3k dataset.

@@

\caption{\label{fig-nldervisPBMCUMAP}(a) Model generated with 21 and 28
bins along x and y axes with hexagonal size 0.03 in the 2D space
overlaid on UMAP data, and (b) high-D model error in model space.}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_1.png}

\subcaption{\label{fig-pbmc1_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_2.png}

\subcaption{\label{fig-pbmc1_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_3.png}

\subcaption{\label{fig-pbmc1_sc3}}
\end{minipage}%

\caption{\label{fig-pbmc1_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/Gnhh4hfsbbg}).}

To find the best NLDR representation for PBMC3k dataset, we observe the
absolute error for different number of non-empty bins and select the one
with the lowest error. As shown in Figure~\ref{fig-abserrorPBMC}, tSNE
with a perplexity set to \(51\), has the lowest error. Therefore, tSNE
with a perplexity of \(51\) is used as the best representation for
PBMC3k dataset.

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-abserrorPBMC-1.pdf}

\caption{\label{fig-abserrorPBMC}Absolute error from UMAP and tSNE
applied to training PBMC3k dataset with diffferent parameter choices.
What is the best parameter choice to create the model? The residual plot
have a steep slope at the beginning, indicating that a smaller number of
non-empty bins causes a larger amount of error. Then, the slope
gradually declines or level off, indicating that a higher number of
non-empty bins generates a smaller error. Using the elbow method, it was
observed that when the number of non-empty bins is set to 439, the
lowest error occurred with the parameters perplexity: 51.}

As shown in Figure~\ref{fig-tsnesuggest}, there are three well-separated
clusters, although they are located close to each other. Additionally,
nonlinear structures can be observed within the clusters (see
Figure~\ref{fig-tsnesuggest}). Hence, the data structure that UMAP
failed to capture for the PBMC3k dataset is successfully captured by
tSNE.

{\centering \includegraphics{paper_files/figure-pdf/fig-tsnesuggest-1.pdf}

\caption{\label{fig-tsnesuggest}2D layout from tSNE (perplexity = 51)
applied for the PBMC3k dataset. Is this a best representation of the
original data?}

Next, the model is fitted for tSNE, and the resulting model is
visualized in the data space. The model shows some quirks, as shown in
Figure~\ref{fig-pbmc2_sc}. The large cluster is divided into three
subclusters, which are in close to each other.

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCtSNE-1.pdf}

\caption{\label{fig-nldervisPBMCtSNE}(a) Model generated with 31 and 40
bins along x and y axes with hexagonal size 0.02 in the 2D space
overlaid on tSNE data, and (b) high-D model error in model space.}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_4.png}

\subcaption{\label{fig-pbmc2_sc4}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_5.png}

\subcaption{\label{fig-pbmc2_sc5}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/pbmc3k/sc_6.png}

\subcaption{\label{fig-pbmc2_sc6}}
\end{minipage}%

\caption{\label{fig-pbmc2_sc}Screen shots of the \textbf{langevitour} of
the PBMC3k data set, shows the model in high-D, a video of the tour
animation is available at (\url{https://youtu.be/R0EroJcGbas}).}

@@

The MNIST dataset consists of grayscale images of handwritten digits
\citep{lecun2010}. \citet{Yingfan2021} used this dataset to demonstrate
how the PaCMAP preserves local structure. We selected the 2D embedding
of PaCMAP for the handwritten digit 1. As shown in
Figure~\ref{fig-pacmapauthor}, we observe a nonlinear structure present
in 2D space. Furthermore, the angle of digit 1 images varies along the
nonlinear structure (see Figure~\ref{fig-sampleimg}). To answer the
following questions,

@@

According to Figure~\ref{fig-mnist1_sc1}, the nonlinear continuity
structure observed in the 2D representation of PaCMAP is also visible
when visualizing the model overlaid on the data space. This provides
evidence that the model accurately captures the structure of the high-D
data. Also, the model has a twisted pattern within the nonlinear
structure, which is an addition pattern of data that cannot be seen in
the 2D representation (see Figure~\ref{fig-mnist1_sc2}).

In order to visually assess the preservation of local structure, it's
important to consider the edges in the model. In a model that
effectively preserves local structure, there shouldn't be any long edges
in high-D space. However, as shown in Figure~\ref{fig-mnist1_sc3}, some
long edges exist in high-D space that are not identified as long edges
in the 2D representation.

@@

\caption{\label{fig-sampleimg}Images of the handwritten digit 1 are
distributed along the right-to-left of first embedding axis of the
PaCMAP nonlinear structure.}

@@

\caption{\label{fig-nldervisMNISTPACMAP}(a) Model generated with 12 and
15 bins along x and y axes with hexagonal size 0.06 in the 2D space
overlaid on PaCMAP data, and (b) high-D model error in model space.}

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/mnist/sc_1.png}

\subcaption{\label{fig-mnist1_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/mnist/sc_2.png}

\subcaption{\label{fig-mnist1_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/mnist/sc_3.png}

\subcaption{\label{fig-mnist1_sc3}}
\end{minipage}%

\caption{\label{fig-mnist1_sc}Screen shots of the \textbf{langevitour}
of the MNIST digit 1 data set, shows the model-in-data space, a video of
the tour animation is available at
(\url{https://youtu.be/zq2GM9qvUNA}).}

In terms of model error, some data points show high error (see
Figure~\ref{fig-nldervisMNISTPACMAP}) due to their deviation from the
typical high-D data structure, acting as outliers. This is evident when
comparing the images associated with these high model error points (see
Figure~\ref{fig-sampleimgerror}) to others (see
Figure~\ref{fig-sampleimg}).

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-sampleimgerror-1.pdf}

\caption{\label{fig-sampleimgerror}Images of handwritten digit 1 which
occur large model error.}
